Universal Properties in Generalized Multicategories

### 2018-01-31 upgm.tex
right modules: problem(!) we need vertical and horizontal heteros

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newcommand{\pto}{\nrightarrow}
\newcommand{\pfrom}{\nleftarrow}
\newcommand{\vcat}{\mathcal}
\newcommand{\cat}{\mathbb}

@@

>>>>

\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newcommand{\pto}{\nrightarrow}
\newcommand{\pfrom}{\nleftarrow}
\newcommand{\vcat}{\mathcal}
\newcommand{\cat}{\mathbb}
\newcommand{\vtkmnd}{\mathbb{K}\text{Mod}(\vcat{V},T)}
\newcommand{\rmod}{\text{RMod}}

@@

A right universal property is a specification for a parameterized
object by defining what maps \emph{into} the object look like.
%
These are simpler to formalize for generalized multicategories because
we have less interaction with the context as given by the monad $T$.
%
If our generalized multicategories are defined as kleisli monoids in
$\vcat V$ with monad $T$, then a right universal property on a
``multicategory'' $\vcat M \in \vtkmnd$ will be parameterized by a
``category'' $\cat C \in \vcat V$ and is defined as an element of the
following profunctor:

\begin{definition}{Profunctor of Right Modules}
  The profunctor of category-parameterized right modules on
  multicategories $\rmod : \vtkmnd \pfrom \vcat V$ has as elements
  $\vcat M \pfrom \cat C$ \emph{right modules} which consist of
  \begin{enumerate}
  \item A profunctor $Q : T \vcat M \pfrom \cat C$ in $\vcat V$.
  \item An action
    \[
\begin{tikzcd}
T^2M_0 \arrow[dd, "\mu"] & T M_0 \arrow[l, "T(M_1)"] & C \arrow[dd, "id"] \arrow[l, "Q"] \\
 & Qcut &  \\
TM_0 &  & C \arrow[ll, "Q"]
\end{tikzcd}
\]
  \item That is associative:
    \[
    \begin{tikzcd}
T^3M_0 \arrow[dd, "\mu"] &  & T^2M_0 \arrow[dd, "\mu"] \arrow[ll, "T^2(M_1)"] & T M_0 \arrow[l, "T(M_1)"] & C \arrow[dd, "id"] \arrow[l, "Q"] &  & T^3M_0 \arrow[dd, "\mu"] & T^2M_0 \arrow[l] & TM_0 \arrow[l] \arrow[dd, "id"] & C \arrow[l, "Q"] \arrow[dd, "id"] \\
 & \mu &  & Qcut &  &  &  & T(M_1) &  &  \\
T^2M_0 \arrow[dd, "\mu"] &  & TM_0 \arrow[ll, "T(M_1)"] &  & C \arrow[ll, "Q"] \arrow[dd, "id"] & = & T^2M_0 \arrow[dd, "\mu"] &  & TM_0 \arrow[ll, "T(M_1)"] & C \arrow[l, "Q"] \arrow[dd, "id"] \\
 &  & Qcut &  &  &  &  &  & Qcut &  \\
TM_0 &  &  &  & C \arrow[llll, "Q"] &  & TM_0 &  &  & C \arrow[lll, "Q"]
\end{tikzcd}
    \]
  \item And unital:
    \[\begin{tikzcd}
 & TM_0 \arrow[ld, "T\eta"] \arrow[d, "id"] &  & C \arrow[ll, "Q"] \arrow[d, "id"] &  &  &  \\
T^2M_0 \arrow[dd, "\mu"] & TM_0 \arrow[l, "M_1"] &  & C \arrow[ll, "Q"] \arrow[dd, "id"] & = & TM_0 \arrow[d, "id"] & C \arrow[d, "id"] \arrow[l, "Q"] \\
 &  & Qcut &  &  & TM_0 & C \arrow[l, "Q"] \\
TM_0 &  &  & C \arrow[lll, "Q"] &  &  & 
    \end{tikzcd}\]
  \end{enumerate}

  And a 2-cell with shape 
  \[\begin{tikzcd}
{\vcat M}_n \arrow[dd, "f"] & \cdots \arrow[l, "R_n"] & {\vcat M}_0 \arrow[l, "R_1"] & {\cat C}_0 \arrow[l, "Q"] & \cdots \arrow[l, "P_1"] &  {\cat C}_m \arrow[l, "P_m"] \arrow[dd, "g"] \\
 &  &  &  &  &  \\
{\vcat M}' &  &  &  &  & {\cat C}' \arrow[lllll, "Q'"]
  \end{tikzcd}\]
  is given by a 2-cell in $\vcat V$ with shape:
  \[\begin{tikzcd}
T^{n+1}\vcat{M}_n \arrow[d, "\mu^n"] & \cdots \arrow[l, "T^nR_n"] & T\vcat{M}_0 \arrow[l, "TR_1"] & \cat C_0 \arrow[l, "Q"] & \cdots \arrow[l, "P_1"] & \cat C_m \arrow[l, "P_m"] \arrow[dd, "g"] \\
T\vcat{M}_n \arrow[d, "Tf"] &  &  &  &  &  \\
T\vcat{M}' &  &  &  &  & \cat C' \arrow[lllll, "Q'"]
\end{tikzcd}
  \]

  with composition defined in the obvious way and unitality and
  associativity properties of 2-cells following from the unitality and
  associativity properties of the actions.
\end{definition}

\subsection{Examples}

The most basic is the identity module which is given by $M_1 : TM_0
\pfrom M_0$ itself. We will use this for representability next.

If $\vcat V$ is cartesian or even virtually cartesian, we can define
the \emph{product} property as $\Delta^{*}(M_1\times M_1) : TM_0
\pfrom M_0\times M_0$ and the \emph{cartesian unit} as $!^*(Id_{1}) :
TM_0 \pfrom 1$.

We will need some technology from the next section to define a general
notion of \emph{closed} multicategory, but we can define the property
of $A \to B$ concretely in many different multicategories.

\subsection{Representability}

As noted above, the hom of a generalized multicategory can be seen as
a module $M_1 : T M_0 \pfrom M_0$ and we can make this into a
parameterized element of $\rmod$ of shape:
  \[
  \begin{tikzcd}
 & \vtkmnd_v \arrow[ldd,"id"] \arrow[rdd, "-_{0}"] &  \\
 & -_1 &  \\
\vtkmnd &  & \vcat V \arrow[ll, "\rmod"]
\end{tikzcd}
  \]
and then we can define a representable negative universal property as
just a representable right module with respect to $-_1$.

In more detail a representation of $Q : \vcat M \pfrom \cat C$ is a
2-cell:
  \[
  \begin{tikzcd}
TM_0 \arrow[dd, "id"] &  & C \arrow[dd, "G_Q"] \arrow[ll, "Q"] \\
 & I_Q &  \\
TM_0 &  & M_1 \arrow[ll, "M_1"]
  \end{tikzcd}
  \]
that is cartesian in the appropriate sense.
%
Returning to type-theoretic intuition, this square is the
\emph{introduction rule}: above the line we have the ``formal''
product or functin etc, and below the line we have the
``internalized'' notion.
%
The associativity rule gives the definition of substitution for the
introductin rule.
%
We can define an ``elimination'' rule using the cartesian property of
$I_Q$, except \emph{we can't we need a notion of vertical morphism from cat to multicat(!)}:

\[\begin{tikzcd}
C \arrow[d, "F"] &  & C \arrow[ll, "\to_C"] \arrow[dd, "id"] &  & C \arrow[dd, "F"] &  & C \arrow[dd, "F"] \arrow[ll, "=_C"] \\
M_0 \arrow[d, "\eta"] & \epsilon &  &  &  & F &  \\
T(M_0) \arrow[dd, "id", two heads, tail] &  & C \arrow[ll, "|-^P"] \arrow[dd, "F"] & = & M_0 \arrow[dd, "\eta"] &  & M_0 \arrow[dd, "id"] \arrow[ll, "=_{M_0}"] \\
 & \iota &  &  &  & Var &  \\
T(M_0) &  & M_0 \arrow[ll, "\vdash"'] &  & T(M_0) &  & M_0 \arrow[ll, "|-"]
\end{tikzcd}
\]
